Review Paper Draft

### 2020-07-01 main.tex
Introduction edits, notational tweaks

Genetic variants are often correlated with each other, meaning that univariate testing of genetic variants may result in false positives and biased effect estimates. This occurs when the variant being tested is correlated with other variants, so that the effects of these correlated variants are erroneously attributed to the variant in question. \anna{One method for analyzing data in the presence of correlated loci, or linkage disequilibrium (LD), is LD-pruning. In LD-pruning methods, SNPs are sequentially thinned, so that only SNPs correlated with each other below some desired threshold are used in modeling and testing procedures.} LD-pruning limits analysis to SNPs which are representative of a particular region. This may result in causal SNPs being removed from analysis, so that only SNPs in LD with the causal loci can be identified and their effects estimated. In addition, it has been shown that in the worst-case scenario, pruning algorithms may not select any representative SNPs in a particular region of the genome \citep{prive2018efficient}. However, LD-pruning does allow for the use of methods which assume uncorrelated loci such as \anna{BLAHHHHH: \url{https://www.nature.com/articles/srep27644/tables/1}}. In addition, LD-pruning may result in improved predictive accuracy \anna{because why}, which can be of particular benefit to polygenic risk score methods.

@@

Lasso penalized regression models typically assume independent observations and uncorrelated errors. In the presence of population structure, some samples are more correlated than others, violating the independent errors assumption. If unaccounted for, this dependency can result in the identification of spurious associations. Though methods to correct for population stratification and cryptic relatedness have been extensively investigated \citep{Amin2007, hoffman2013correcting, price2006principal, Rakitsch2012, bhatnagar2019simultaneous, Sillanpaeae2011}, there remain open and relevant problems \pb{this is vague} in analyzing structured genetic data \citep{lawson2019population, barton2019population}.

@@

%------------------------------%
\subsection{PC-Lasso}
PC-Lasso uses the first $k$ principal components of $\bX$ to approximate $\bZ\bgamma$ and includes them in the model as unpenalized covariates. It has been shown \citep{hoffman2013correcting} that this corresponds to the regression model
\begin{equation}
    \label{eqn:pca_reg}
    \by = \Int + \bX\bbeta + \bU_{1:k}\boldsymbol{\alpha} + \beps 
\end{equation}
where $\bU_{1:k}$ is the $n \times k$ matrix of principal component vectors, $\boldsymbol{\alpha}$ their corresponding $k \times 1$ coefficient vector, and all other parameters are as defined in equation \eqref{eqn:model}. In a penalized regression framework, this model yields the objective function
\begin{equation}
    \label{eqn:pca_obj}
    Q(\boldsymbol{\beta, \alpha| X, y}) = \frac{1}{2n} ||\by - \Int - \bX \bbeta- \bU_{1:k}\boldsymbol{\alpha}||_2^2 + \lambda || \bbeta ||_1
\end{equation}
where $\lambda$ is a tuning parameter, and $|| \boldsymbol{\beta} ||_1 = \sum_{j=1}^p |\beta_j|$ denotes the $\ell_1$ norm of the regression coefficients, and $||.||_2$ the $\ell_2$ norm. Note that the principal component regression coefficients, $\boldsymbol{\alpha}$, are not included in the penalty term. Modeling unobserved confounding using fixed effects in this manner thus requires the estimation of $p + k + 1$ parameters.

%------------------------------%
\subsection{LMM-Lasso}
LMM-Lasso models $\bZ\bgamma$ as a random effect assuming 
$$\bgamma \sim N\left(\boldsymbol{0}, \frac{\sigmagg}{p} \bI_p \right).$$  Typically, $\bZ$ is not observed directly, and a common convention is to assume $\bZ = \bX$ \citep{wang2018multiplex, lippert2011fast, yang2014advantages}.This results in the LMM regression model
\begin{equation}
    \label{eqn:lmm_reg}
    \by = \Int + \bX\bbeta + \bX\bgamma+ \beps,
\end{equation}
where we assume $\beps \independent \bX\bgamma$. Rather than assuming anything about $\bgamma$ directly, an equivalent representation lets $\bu=\bX\bgamma$ such that $\bu \sim N(\boldsymbol{0}, \sigmagg \bK)$, and yields the equivalent LMM model
\begin{equation}
    \label{eqn:lmm_reg_u}
    \by = \Int + \bX\bbeta + \bu + \beps.
\end{equation}
Regardless of representation, the LMM assumes $\by \sim N(\Int + \bX \bbeta, \bV)$, where $\bV = \sigmagg \textbf{K} + \sigmaee \textbf{I}$. LMM-Lasso performs a weighted rotation of the original data, left-multiplying by $\bV^{-1/2}$ in a manner akin to generalized least squares to diagonalize the error term such that
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \left(\Int +\bX \bbeta\right), \bI).
\end{equation}
\anna{...however, as Karl Rohe paper shows, GLS does not completely translate to penalized regression...comment on this?}
\begin{comment}
For ease of notation, we incorporate the intercept into the design matrix and let $\bX' = \begin{bmatrix} \textbf{1} & \bX \end{bmatrix}$ and $\bbeta' = \begin{pmatrix} \mu & \bbeta \end{pmatrix}^T$ so that we can express this more compactly as 
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \bX' \bbeta', \bI).
\end{equation}
\end{comment}
Using the spectral decomposition of $\bK$ described in equation \eqref{eqn:k}, we can gain further insight into the mechanism of this transformation. Note that $\bV^{-1}$ can be expressed using the following matrix factorization 
\begin{align*}
    \bV^{-1} &= (\sigmagg \bK + \sigmaee \bI)^{-1}\\
    &=(\sigmagg \bU \bD^2 \bU^T + \sigmaee \bU \bU^T)^{-1}\\
    &= (\bU (\sigmagg \bD^2 + \sigmaee \bI) \bU^T)^{-1}\\
    &=\bU (\sigmagg \bD^2 + \sigmaee \bI)^{-1} \bU^T.
\end{align*}
By incorporating these decompositions, the transformed data can be expressed as $\bW \bXT$ and $\bW \byT$, respectively, where $\bXT = \bU^T \bX$ and $\byT = \bU^T \left(\by - \Int \right)$, where we center the transformed $\by$ to simplify notation and computation.  $\bW = \text{diag}\{w_1, ... w_n\}$ where $w_i$ corresponds to the square root of the $i$th element of the diagonal matrix $(\sigmagg \boldsymbol{D}^2 + \sigmaee \bI)^{-1}$. Together, this yields the objective function
\begin{equation}
\label{eqn:lmm_obj}
Q(\bbeta | \bXT, \byT, \bW) = || \bW (\byT - \bXT \bbeta)||_2^2 + \lambda || \bbeta ||_1.
\end{equation}
An advantage to modeling unobserved confounding using a random effect, as opposed to fixed effects as in PC-Lasso, is that this requires the estimation of substantially fewer parameters: $p + 2$ for LMM-Lasso-Rakitsch, and $p + 3$ for LMM-Lasso-ggmix.

@@

In carrying out simulation studies of the LMM regression model \eqref{eqn:lmm_reg}, a common practice \citep{Rakitsch2012, bhatnagar2019simultaneous} is to draw a new value of the quantity $\bX\bgamma$ from a $N(\boldsymbol{0}, \sigmagg \bK)$ for each replication of the analysis.  We argue that this is somewhat misleading from the perspective of understanding the statistical properties of the method. Averaging over situations in which the environmental confounding is constantly changing direction and magnitude ``washes out'' bias in $\hat{\bbeta}$ while inflating the variance. To see why this is the case, consider the low-dimensional ordinary least squares (OLS) estimates of $\bbeta$ based on model \eqref{eqn:lmm_reg} where, without loss of generatlity, we assume $\mu = 0$ for ease of notation: \anna{intercept stuff would get messy here... does this work?}
\begin{align*}
    \mathbb{E}(\hat{\bbeta}) &= (\bX^T \bX)^{-1} \bX^T \mathbb{E}(\by)\\
    &=  (\bX^T \bX)^{-1} \bX^T \mathbb{E}(\bX\bbeta + \bX\bgamma + \beps)\\
    &\quad\\
    \mathbb{V}(\hat{\bbeta}) &= (\bX^T \bX)^{-1} \bX^T \mathbb{V}(\by)\\
    &=  (\bX^T \bX)^{-1} \bX^T \mathbb{V}(\bX\bbeta + \bX\bgamma + \beps)\\
    &=  (\bX^T \bX)^{-1} \bX^T \mathbb{V}(\bX\bgamma + \beps) \bX  (\bX^T \bX)^{-1}.
\end{align*}
The assumption that environmental effects follow a random distribution, $\bgamma \sim N(\boldsymbol{0}, \sigmagg / p \bI_p)$, yields $\mathbb{E}(\hat{\bbeta}) = \bbeta$, and $\mathbb{V}(\hat{\bbeta}) = \sigmagg/p \bI + \sigmaee (\bX^T \bX)^{-1}$. Returning to our earlier example involving confounding due to air quality, randomly drawing a new value of $\bX\bgamma$ for each replication would mean that subpopulation A may be exposed to good air quality in one replication, and bad air quality in the next. Reporting the average across these situations is inconsistent with the question we are truly interested in, namely: what would happen if we took repeated samples from fixed populations A and B? Those repeated samples would show a systematic bias due to the difference in air quality, but this is masked if we generate new values of $\bX\bgamma$ with each replication. Considering $\bgamma$ as a fixed quantity is more consistent with the usual meaning of environmental confounding. Specifically, in the low-dimensional OLS setting, under the assumption that the quantity $\bgamma$ is fixed, $\mathbb{E}(\hat{\bbeta}) = \bbeta + \bgamma, \mathbb{V}(\hat{\bbeta}) = \sigmaee (\bX^T \bX)^{-1}$, which allows us to quantify bias in our estimates.

@@

Define the projection matrix of $\bX$ by $\boldsymbol{P}_{\bX} = \bX (\bX^T \bX)^{-1} \bX^T$, and let $\mathcal{C}(\boldsymbol{P}_{\bX})$ and $\mathcal{N}(\boldsymbol{P}_{\bX})$ denote the column and null spaces of $\boldsymbol{P}_{\bX}$, respectively. Then, $\exists \boldsymbol{\tau, \psi}: \bZ \bgamma = \bX \boldsymbol{\tau + \psi}$ , $\boldsymbol{\tau} \in \mathcal{C}(\boldsymbol{P}_{\bX})$, and $\boldsymbol{\psi} \in \mathcal{N}(\boldsymbol{P}_{\bX})$ such that
\begin{align*}
    \by &= \bX\bbeta + \bZ\bgamma + \beps\\
    &= \bX\bbeta + \bX \boldsymbol{\tau} + \boldsymbol{\psi} + \beps\\
    &=  \bX \bbeta' + \beps'
\end{align*}
where $\boldsymbol{\tau}$ quantifies bias in the estimated $\boldsymbol{\hat{\beta}'}$ and $\boldsymbol{\psi}$ is absorbed by the error term. In other words, any model specified in terms of $\bZ$ and $\bgamma$ is equivalent to a parameterization using $\bX$ and $\boldsymbol{\tau}$.

@@

\pb{Just FYI, it's hard to tell bias and var apart in black and white} \anna{will fix this and greek letters}

>>>>

Genetic variants are often correlated with each other, meaning that univariate testing of genetic variants may result in false positives and biased effect estimates. This occurs when the variant being tested is correlated with other variants, so that the effects of these correlated variants are erroneously attributed to the variant in question. One method for analyzing data in the presence of correlated loci, or linkage disequilibrium (LD), is LD pruning. LD pruning can be an important preprocessing step in a GWAS analysis, for which there exist a variety of algorithms and methods. The overarching approach of these methods is to sequentially thin SNPs, so that only SNPs correlated with each other below some desired threshold are used in modeling and testing procedures. This may result in causal SNPs being removed from analysis, so that only SNPs in LD with the causal loci, as opposed to the causal loci themselves, can be identified and their effects estimated.

% LD pruning thus limits analysis to SNPs which are representative of a particular region.

% In some scenarios pruning algorithms may not select any representative SNPs in a particular region of the genome \citep{prive2018efficient}. However, LD-pruning has been shown to improve performance of principal component analysis (PCA) methods \citep{abdellaoui2013population} and \anna{what} \anna{because why}.

@@

Lasso penalized regression models typically assume independent observations and uncorrelated errors. In the presence of population structure, some samples are more correlated than others, violating the independent errors assumption. If unaccounted for, this dependency can result in the identification of spurious associations. Though methods to correct for population stratification and cryptic relatedness have been extensively investigated \citep{Amin2007, hoffman2013correcting, price2006principal, Rakitsch2012, bhatnagar2019simultaneous, Sillanpaeae2011}, until recently the majority of this work has focused on plant and animal breeding data, much of which experimentally controlled. While many standard correction techniques may in principal be applied to human genetic data, their performance must be studied within the context of human ancestry as well as non-experimental conditions \citep{lawson2019population, barton2019population}.

@@

%------------------------------%
\subsection{PC-Lasso}
PC-Lasso uses the first $k$ principal components of $\bX$ to approximate $\bZ\bgamma$ and includes them in the model as unpenalized covariates. It has been shown \citep{hoffman2013correcting} that this corresponds to the regression model
\begin{equation}
    \label{eqn:pca_reg}
    \by = \Int + \bX\bbeta + \bU_{1:k}\boldsymbol{\alpha} + \beps 
\end{equation}
where $\bU_{1:k}$ is the $n \times k$ matrix of principal component vectors, $\boldsymbol{\alpha}$ their corresponding $k \times 1$ coefficient vector, and all other parameters are as defined in equation \eqref{eqn:model}. In a penalized regression framework, this model yields the objective function
\begin{equation}
    \label{eqn:pca_obj}
    Q(\bbeta, \boldsymbol{\alpha}|\bX, \by) = \frac{1}{2n} ||\by - \Int - \bX \bbeta- \bU_{1:k}\boldsymbol{\alpha}||_2^2 + \lambda || \bbeta ||_1
\end{equation}
where $\lambda$ is a tuning parameter, and $|| \boldsymbol{\beta} ||_1 = \sum_{j=1}^p |\beta_j|$ denotes the $\ell_1$ norm of the regression coefficients, and $||.||_2$ the $\ell_2$ norm. Note that the principal component regression coefficients, $\boldsymbol{\alpha}$, are not included in the penalty term. Modeling unobserved confounding using fixed effects in this manner thus requires the estimation of $p + k + 1$ parameters.

%------------------------------%
\subsection{LMM-Lasso}
LMM-Lasso models $\bZ\bgamma$ as a random effect assuming 
$$\bgamma \sim N\left(\boldsymbol{0}, \frac{\sigmagg}{p} \bI_p \right).$$  Typically, $\bZ$ is not observed directly, and a common convention is to assume $\bZ = \bX$ \citep{wang2018multiplex, lippert2011fast, yang2014advantages}.This results in the LMM regression model
\begin{equation}
    \label{eqn:lmm_reg}
    \by = \Int + \bX\bbeta + \bX\bgamma+ \beps,
\end{equation}
where we assume $\beps \independent \bX\bgamma$. Rather than assuming anything about $\bgamma$ directly, an equivalent representation lets $\bu=\bX\bgamma$ such that $\bu \sim N(\boldsymbol{0}, \sigmagg \bK)$, and yields the equivalent model
\begin{equation}
    \label{eqn:lmm_reg_u}
    \by = \Int + \bX\bbeta + \bu + \beps.
\end{equation}
Regardless of representation, the LMM assumes $\by \sim N(\Int + \bX \bbeta, \bV)$, where $\bV = \sigmagg \textbf{K} + \sigmaee \textbf{I}$. LMM-Lasso performs a weighted rotation of the original data, left-multiplying by $\bV^{-1/2}$ in a manner akin to generalized least squares to diagonalize the error term such that
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \left(\Int +\bX \bbeta\right), \bI).
\end{equation}
\anna{...however, as Karl Rohe paper shows, GLS does not completely translate to penalized regression...comment on this?}
\begin{comment}
For ease of notation, we incorporate the intercept into the design matrix and let $\bX' = \begin{bmatrix} \textbf{1} & \bX \end{bmatrix}$ and $\bbeta' = \begin{pmatrix} \mu & \bbeta \end{pmatrix}^T$ so that we can express this more compactly as 
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \bX' \bbeta', \bI).
\end{equation}
\end{comment}
Using the spectral decomposition of $\bK$ described in equation \eqref{eqn:k}, we can gain further insight into the mechanism of this transformation. Note that $\bV^{-1}$ can be expressed using the following matrix factorization 
\begin{align*}
    \bV^{-1} &= (\sigmagg \bK + \sigmaee \bI)^{-1}\\
    &=(\sigmagg \bU \bD^2 \bU^T + \sigmaee \bU \bU^T)^{-1}\\
    &= (\bU (\sigmagg \bD^2 + \sigmaee \bI) \bU^T)^{-1}\\
    &=\bU (\sigmagg \bD^2 + \sigmaee \bI)^{-1} \bU^T.
\end{align*}
By incorporating these decompositions, the transformed data can be expressed as $\bW \bXT$ and $\bW \byT$, respectively, where $\bXT = \bU^T \bX$ and $\byT = \bU^T \left(\by - \Int \right)$, where we center the transformed $\by$ to simplify notation and computation \anna{does this make sense?}.  $\bW = \text{diag}\{w_1, ... w_n\}$ where $w_i$ corresponds to the square root of the $i$th element of the diagonal matrix $(\sigmagg \boldsymbol{D}^2 + \sigmaee \bI)^{-1}$. Together, this yields the objective function
\begin{equation}
\label{eqn:lmm_obj}
Q(\bbeta | \bXT, \byT, \bW) = || \bW (\byT - \bXT \bbeta)||_2^2 + \lambda || \bbeta ||_1.
\end{equation}
An advantage to modeling unobserved confounding using a random effect, as opposed to fixed effects as in PC-Lasso, is that this requires the estimation of substantially fewer parameters: $p + 2$ for LMM-Lasso-Rakitsch, and $p + 3$ for LMM-Lasso-ggmix.

@@

In carrying out simulation studies of the LMM regression model \eqref{eqn:lmm_reg}, a common practice \citep{Rakitsch2012, bhatnagar2019simultaneous} is to draw a new value of the quantity $\bX\bgamma$ from a $N(\boldsymbol{0}, \sigmagg \bK)$ for each replication of the analysis.  We argue that this is somewhat misleading from the perspective of understanding the statistical properties of the method. Averaging over situations in which the environmental confounding is constantly changing direction and magnitude ``washes out'' bias in $\hat{\bbeta}$ while inflating the variance. To see why this is the case, consider the low-dimensional ordinary least squares (OLS) estimates of $\bbeta$ based on model \eqref{eqn:lmm_reg} where, without loss of generatlity, we assume $\mu = 0$ for ease of notation: \anna{intercept stuff would get messy here... does this work?}
\begin{align*}
    \mathbb{E}(\hat{\bbeta}) &= (\bX^T \bX)^{-1} \bX^T \mathbb{E}(\by)\\
    &=  (\bX^T \bX)^{-1} \bX^T \mathbb{E}(\bX\bbeta + \bX\bgamma + \beps)\\
    &\quad\\
    \mathbb{V}(\hat{\bbeta}) &= (\bX^T \bX)^{-1} \bX^T \mathbb{V}(\by)\\
    &=  (\bX^T \bX)^{-1} \bX^T \mathbb{V}(\bX\bbeta + \bX\bgamma + \beps)\\
    &=  (\bX^T \bX)^{-1} \bX^T \mathbb{V}(\bX\bgamma + \beps) \bX  (\bX^T \bX)^{-1}.
\end{align*}
The assumption that environmental effects follow a random distribution, $\bgamma \sim N(\boldsymbol{0}, \sigmagg / p \bI_p)$, yields $\mathbb{E}(\bbetaHat) = \bbeta$, and $\mathbb{V}(\bbetaHat) = \sigmagg/p \bI + \sigmaee (\bX^T \bX)^{-1}$. Returning to our earlier example involving confounding due to air quality, randomly drawing a new value of $\bX\bgamma$ for each replication would mean that subpopulation A may be exposed to good air quality in one replication, and bad air quality in the next. Reporting the average across these situations is inconsistent with the question we are truly interested in, namely: what would happen if we took repeated samples from fixed populations A and B? Those repeated samples would show a systematic bias due to the difference in air quality, but this is masked if we generate new values of $\bX\bgamma$ with each replication. Considering $\bgamma$ as a fixed quantity is more consistent with the usual meaning of environmental confounding. Specifically, in the low-dimensional OLS setting, under the assumption that the quantity $\bgamma$ is fixed, $\mathbb{E}(\bbetaHat) = \bbeta + \bgamma, \mathbb{V}(\bbetaHat) = \sigmaee (\bX^T \bX)^{-1}$, which allows us to quantify bias in our estimates.

@@

Define the projection matrix of $\bX$ by $\boldsymbol{P}_{\bX} = \bX (\bX^T \bX)^{-1} \bX^T$, and let $\mathcal{C}(\boldsymbol{P}_{\bX})$ and $\mathcal{N}(\boldsymbol{P}_{\bX})$ denote the column and null spaces of $\boldsymbol{P}_{\bX}$, respectively. Then, $\exists \boldsymbol{\tau, \psi}: \bZ \bgamma = \bX \boldsymbol{\tau + \psi}$ , $\boldsymbol{\tau} \in \mathcal{C}(\boldsymbol{P}_{\bX})$, and $\boldsymbol{\psi} \in \mathcal{N}(\boldsymbol{P}_{\bX})$ such that
\begin{align*}
    \by &= \bX\bbeta + \bZ\bgamma + \beps\\
    &= \bX\bbeta + \bX \boldsymbol{\tau} + \boldsymbol{\psi} + \beps\\
    &=  \bX \bbeta' + \beps'
\end{align*}
where $\boldsymbol{\tau}$ quantifies bias in the estimated $\bbetaHat'$ and $\boldsymbol{\psi}$ is absorbed by the error term. In other words, any model specified in terms of $\bZ$ and $\bgamma$ is equivalent to a parameterization using $\bX$ and $\boldsymbol{\tau}$.

@@

\pb{Just FYI, it's hard to tell bias and var apart in black and white} \anna{will fix this, greek letters, labels}
